Grove: A Convergent Collaborative Structure-Editor Calculus

### 2022-08-18 introduction.tex
intro

Programming is an increasingly collaborative activity.
Today, collaborative editing services are either provided directly by an asynchronous,
diff-and-merge-based version control system like git, darcs, or pijul,
or they are integrated into a real-time editor built on top of a distributed consistency protocol such as OT or CRDT.

Diff-and-merge is not an ideal collaborative editing solution.
Since diffs do not provide information on the original scope or ordering of edits,
the merging process must reconstruct an approximation, e.g., as a sequence of line additions and deletions.
Clearly, in the context of editing programs,
a line-based merge is not only less likely to succeed than one which retains finer grained information about changes over time,
it is also less likely to produce useful information about conflicting edit states, when they arise.

Operational Transformation (OT), on the other hand, goes perhaps too far in the opposite direction.
OT is a distributed system architecture that coordinates the proliferation and maintainance of edit states and actions.
Industrial-scale collaborative editors such as Google Docs tend to employ OT
as it provides strong theoretical guarantees on the eventual consistency of causally-ordered events
while remaining suitable for real-time applications.
It does so by maintaining a history of recent edits, or operations, while transforming individual edits to preserve a causal ordering.
But fully featured distributed systems that maintain causal orderings of arbitrary events are hard to design well,
so in practice OT-based systems tend to be restricted to linear text editing or have fewer features than comparable single-user alternatives.

Convergent Replicated Data Types (CRDTs) were proposed as a reaction to the complexity of Operational Transformation,
though the claim is currently under strutiny by the research community.
A CRDT is essentially a distributed data structure with an API for building and interacting with it.
CRDTs avoid the need for explicit state synchronization by requiring that allowable edit actions inhabit a lattice.

@@

% start with something that everyone can accept: ``programming is collaborative.''

% briefly introduce 3 types: diff, OT, CRDT

@@

% general motivation: programming is a collaborative activity

% today usually takes the form of coarse-grained solutions, e.g, git

% if you try to use eg google docs for this, there are several problems
% (gap problem, diffs don't handle repositioning well)

% diff is not really an ideal way of communicating change: e.g., how repositioning gets mishandled

% CRDT: tree-native vs text. think about obvious limitation of existing tools

>>>>

Programming is an increasingly collaborative activity.
Today, collaborative editing services are either provided directly by an asynchronous,
diff-and-merge-based version control system like git, darcs, or pijul,
or they are integrated into a real-time editor built on top of a distributed consistency protocol such as OT or CRDT.
Whereas algorithms for collaborative text editing have been well developed,
we know far less about collaborative editing of more structured documents---specifically, of program source files---%
hence we are motivated to ask the following questions:
\begin{itemize}
\item Is there a more natural model for collaborative editing of tree structures?
\item How do we preserve user intent in the presence of concurrent, non-linear edits?
\end{itemize}

Diff-and-merge is clearly not an ideal solution.
Since diffs do not provide information on the original scope or ordering of edits,
the merging process must reconstruct an approximation, e.g., as a sequence of line additions and deletions.
In a tree-native setting,
line-based merging is not only less likely to succeed than one which retains finer grained information,
it is also less able to provide useful information about conflicting edit states as they arise.
Moreover, because diffs cannot guarantee preservation of user intent,
critical features such as identity-preserving repositioning are impossible to implement on top of them.

Operational Transformation (OT), on the other hand, goes perhaps too far in the opposite direction.
It is a distributed system architecture that coordinates the proliferation and maintainance of edits and edit states.
Industrial-scale collaborative editors, such as Google Docs,
employ OT for real-time applications requiring strong eventual consistency of causally ordered events.
It does so by maintaining an event log it transforms upon each new entry in order to preserve the causal ordering.

The Convergent Replicated Data Type (CRDT) was proposed as a simpler alternative to the OT architecture.
CRDTs do not require causal ordering maintenance because the language of edits must form a join semilattice,
thereby ensuring the edit state update function increases monotoncally.
(Foreshadow solution by warning about a shortcoming of CRDT.)

(Propose Grove as a solution)

% OT is a natural fit for character-based editing, but the architecture is hard to extend, so it is rarely used for more structured documents.
% CRDT imposes less restrictions on the structure of the underlying document,
% providing a more natural substrate for collaborative structure editing.

@@

@@
